The effect of prenatal environment on brain metabolic function and perception in a lizard

### 2025-10-12 ms.qmd
updates

Pablo Recio^1^,$\ddagger$ , Dalton C. Leibold^1^, Ondi L. Crino^1,2^, Christopher R. Friesen^3,4^, Daniel W.A. Noble^1^

@@

## Abstract
Cognitive processes such as the ability to perceive prey, are crucial for survival and reproduction. However, early-life conditions like stress-related hormones or thermal environments, can shape cognitive abilities by influencing brain structure and function. Mitochondrial physiology is thought to be a central mechanism underlying these effects given that it is crucial for the energy production needed for brain function. Here, we investigated the combined influence of prenatal corticosterone (CORT) – a key stress hormone in reptiles – and incubation temperature on brain mitochondrial function and impacts on prey detection in the delicate skink (_Lampropholis delicata_). We manipulated egg CORT levels and incubation temperature, then assessed the ability of lizards to detect chemical and visual prey stimuli. Using flow cytometry, we measured metabolic function, reactive oxygen species (ROS) production, and oxidative stress in the olfactory bulbs and optic tecta-two brain regions involved in sensory processing. We predicted that CORT exposure and high temperatures would interact to shape perception of prey via changes in mitochondrial respiration and oxidative damage. While metabolic function remained robust to early life conditions, CORT and temperature interacted to influence oxidative stress in the optic tecta but not the olfactory bulbs. Additionally, CORT-treated lizards responded faster when presented with chemical cues but not visual stimuli. Our findings suggest that prenatal conditions can have lasting effects on brain physiology and cognition, though the role of mitochondrial function in these processes remains unclear. Future research should explore alternative mechanisms linking early-life stress, cell metabolism, and cognitive performance.

@@

Cognition encompasses the ways in which animals acquire, process, and store information, enabling perception, learning, memory, and decision-making [@shettleworth]. It is essential for survival and reproduction, allowing individuals to adapt to changing environments [@dukas_evolutionary_2004]. However, cognitive abilities can vary considerably between individuals, with differences arising from genetic factors, environmental conditions, or a combination of both [@sakata_neural_2000; @dukas_evolutionary_2004]. The prenatal environment, in particular, plays a critical role in shaping brain development and cognition across a wide range of species [@sakata_neural_2000; @zhu_prenatal_2004; @crino_corticosterone_2014-learn; @amiel_effects_2017]. However, the impact of early-life environments has primarily been studied in the context of learning and memory [@zhu_prenatal_2004; @crino_corticosterone_2014-learn; @bebus_associative_2016; @abayarathna_effects_2020], neglecting how other equally important cognitive domains might be affected [but see @burger_antipredator_1998; @vila_pouca_quantity_2019]. Perception is an often overlooked, but fundamental skill [@shettleworth]. Perceiving different stimuli is essential for locating food, avoiding predators, or interacting with conspecifics [@burger_antipredator_1998; @desfilis2003stimulus; @recio2023conspecific]. Variation in the ability to detect or recognize relevant stimuli can have significant consequences for fitness [@pyke1984optimal]. For example, visual acuity in the bluegill sunfish (_Lepomis macrochirus_) allow animals to detect prey at large distances, which can increase their foraging efficiency and energy intake [@hairston1982fish]. Therefore, understanding how prenatal conditions affect the ability to detect prey can be important to understanding the broader consequences of early-life environments on fitness.

Prenatal environments influence cognitive abilities because the brain is particularly sensitive to environmental insults during early stages of development [@zhu_prenatal_2004]. Early-life conditions can shape cognition by altering gene expression [@zhou2020effects], neurotransmitter production [@amani2021perinatal], or brain structure and function [@amiel_effects_2017]. Long-lasting effects on energy production and oxidative stress in the brain may be a key mechanism by which cognitive function is impacted by developmental environments [@siegel1994basic; @du_dynamic_2009; @picard_mitochondria_2014; @picard_energetic_2018]. Given the high energetic demands of cognitive processes [@mcnay_decreases_2000; @mann_coupling_2021; @alexandrov_neuronal_2022], cognitive performance is expected to depend on efficient mitochondrial respiration. Additionally, high cognitive abilities are typically associated with increased neuron density and functionality [@lefebvre_taxonomic_2011; @amiel_effects_2017], both of which can be impaired by excessive reactive oxygen species (ROS) production and oxidative stress [@finkel_oxidants_2000; @zhu_prenatal_2004; @du_dynamic_2009; @gong_chronic_2011; @hoffmann_mitochondrion_2018]. For instance, Hara et al. (2014) found that performance in visuospatial working-memory tasks correlate negatively with the number of round-shaped mitochondria-associated with higher oxidative stress-per presynaptic terminal. Findings from different studies suggest that mitocondrial activity can be essential in shaping cognitive abilities. Thus, understanding how early-life conditions affect mitochondrial physiology may provide insights into the mechanisms underlying cognitive performance.

Mitochondria are highly responsive to environmental conditions during early development, with potential short-term and long-lasting consequences [reviewed in @gyllenhammer2020developmental]. For example, environmental cues such as external conditions or hormone exposure can cause immediate disruptions to metabolic function, triggering cascading effects on oxidative stress that may persist later in life [reviewed in @gyllenhammer2020developmental]. Among the range of environmental factors that can shape mitochondrial physiology, some exert particularly strong and consistent effects. Stress-related hormones and temperature are particularly relevant for mitochondrial function due to their strong effects on metabolic processes [@sapolsky_how_2000; @stier2022experimental; @crino2024eggs]. Under stressful situations, animals instigate adaptive physiological and behavioural adjustments that involve increases in glucocorticoids (GCs), which are also important metabolic regulators [@sapolsky_how_2000]. Similarly, temperature plays a prominent role in regulating metabolic activity in ectotherms [@stier2022experimental; @crino2024eggs]. In fact, heightened GC levels or changes in the thermal environment during early development have profound and sustained effects on mitochondrial respiration efficiency or oxidative stress [@zhu_prenatal_2004; @song2009prenatal; @costantini2011meta; @haussmann_embryonic_2012; @treidel2016temperature; @stier2022experimental; @crino2024eggs]. However, how the early environment influences mitochondrial physiology, and its associated cognitive consequences remains largely unknown outside of mammals [see @chaudhari2022early for a review]. In addition, whether and how these interactions affect perceptual abilities, such as prey detection, is still largely unexplored.

@@

#### Statistical analyses
We performed the analyses for each brain region/stimulus and each variable (mitochondrial density, metabolic capacity, metabolic capatity/mitochondrial density, ROS, ROS/mitochondrial density, DNA damage, lipid peroxidation, and detection latency) separately. We first fit a set of preliminary models for all the interest variables where we included the hormone (CORT versus Control), temperature (Cold versus Hot), and their interaction. These models also included the sex and age of the lizards at the time where the trials started (for detection latency) or when the lizards were euthanized (for all mitochondrial-related variables). For detection latency, we also included previous experience with the prey (familiar versus unknown), their performance on the motivation test [if they ate the cricket (1), or not (0)], and the interaction between CORT and motivation as fixed effects. We included the interaction between CORT and motivation because CORT can impact appetite [@conde2018stress]. All models included clutch identity as a random factor to account for clutch effects. For models of detection latency we also included a random effect of lizard identity because of our repeated measures design. The structure and results of these models are provided on Tables S1-S22 in _Supplementary Material_. After the preliminary models, we simplified models by excluding some of the factors that were not significant. However, we always included the main interest factors: CORT, temperature and their interaction. Random factors remain the same for the final models. All the response variables were mean centered and standardized by dividing by two times the standard deviation [@gelman2008scaling]. Before standardization, mitochondrial density, DNA damage, lipid peroxidation, and detection latency were log-transformed.

@@

_Optic tecta_: We did not find significant effects of CORT, temperature, or their interaction on mitochondrial density, metabolic capacity, or ROS production in the optic tecta, neither of CORT or incubation temperature on DNA damage or lipid peroxidation (see @fig-results_energy, @fig-results_oxidative, and Table S2). However, the interaction of CORT and temperature have significant effects on DNA damage ([(β~Control-Hot~ - β~CORT-Hot~) - (β~Control-Cold~ - β~CORT-Cold~)]: mean = `r format_dec((mean(DNA_OT$Control_Hot) - mean(DNA_OT$CORT_Hot)) - (mean(DNA_OT$Control_Cold) - mean(DNA_OT$CORT_Cold)), 3)`, pMCMC `r format_p(pmcmc((DNA_OT$Control_Hot - DNA_OT$CORT_Hot) - (DNA_OT$Control_Cold - DNA_OT$CORT_Cold)), 3, equal = TRUE)`), and lipid peroxidation ([(β~Control-Hot~ - β~CORT-Hot~) - (β~Control-Cold~ - β~CORT-Cold~)]: mean = `r format_dec((mean(LP_OT$Control_Hot) - mean(LP_OT$CORT_Hot)) - (mean(LP_OT$Control_Cold) - mean(LP_OT$CORT_Cold)), 3)`, pMCMC `r format_p(pmcmc((LP_OT$Control_Hot - LP_OT$CORT_Hot) - (LP_OT$Control_Cold - LP_OT$CORT_Cold)), 3, equal = TRUE)`) in the optic tecta. CORT elevations had no effect on hot-incubated lizards (β~Control-Hot~ - β~CORT-Hot~: mean = `r format_dec(mean(DNA_OT$Control_Hot) - mean(DNA_OT$CORT_Hot),3)`, pMCMC `r format_p(pmcmc(DNA_OT$Control_Hot - DNA_OT$CORT_Hot), 3, equal = TRUE)`), but it decreased total DNA damage in cold-incubated animals (β~Control-Cold~ - β~CORT-Cold~: mean = `r format_dec(mean(DNA_OT$Control_Cold) - mean(DNA_OT$CORT_Cold),3)`, pMCMC `r format_p(pmcmc(DNA_OT$Control_Cold - DNA_OT$CORT_Cold), 3, equal = TRUE)`) (see @fig-results_oxidative C, E). The direction of the effects of CORT on lipid peroxidation were the opposite on eggs incubated at high temperatures (β~Control-Hot~ - β~CORT-Hot~: mean = `r format_dec(mean(LP_OT$Control_Hot) - mean(LP_OT$CORT_Hot),3)`, pMCMC `r format_p(pmcmc(LP_OT$Control_Hot - LP_OT$CORT_Hot), 3, equal = TRUE)`) than at low temperatures (β~Control-Cold~ - β~CORT-Cold~: mean = `r format_dec(mean(LP_OT$Control_Cold) - mean(LP_OT$CORT_Cold),3)`, pMCMC `r format_p(pmcmc(LP_OT$Control_Cold - LP_OT$CORT_Cold), 3, equal = TRUE)`), with significant differences between CORT-Hot and CORT-Cold lizards (β~CORT-Hot~ - β~CORT-Cold~: mean = `r format_dec(mean(LP_OT$CORT_Hot) - mean(LP_OT$CORT_Cold),3)`, pMCMC `r format_p(pmcmc(LP_OT$CORT_Hot - LP_OT$CORT_Cold), 3, equal = TRUE)`; see @fig-results_oxidative C). In addition, we found that females had higher DNA damage levels, and lipid peroxidation increased with age (see Table S4).

@@

_Chemical perception (Olfactory bulbs)_: We found a significant effect of mitochondrial density on ROS production (mean β~Mit density - ROS~ = `r format_dec(mean(OB_coeff_mitodensity_ros), 3)`, pMCMC `r format_p(pmcmc(OB_coeff_mitodensity_ros), 3, equal = TRUE)`). However, there were no significant relationships between mitochondrial function or oxidative stress on prey detection, nor between other any other variables (see @fig-sem_results_OB and Table S5).

@@

## Discussion
We examined the impact of prenatal temperature and CORT on visual and chemical perception and determined whether metabolic function underpinned any perceptual differences, as previously shown in mammals [@hara_presynaptic_2014; @chaudhari2022early]. Overall, metabolic processes appeared largely resilient to prenatal environment. However, early-life conditions produced region- and context-dependent effects on oxidative damage and prey detection. We observed interactive effects of prenatal CORT and incubation temperature on lipid proxidation in both regions, while the interaction affected DNA damage only in the optic tecta. Behaviourally, CORT enhanced detection of chemical but not visual prey stimuli. Yet, we found no evidence that metabolism explained behavioural variation. Below, we discuss these findings in the context of mitochondrial metabolic function, oxidative stress, and perception.

#### Metabolic processes in multiple brain regions are robust to early life conditions
Contrary to our predictions, we found no significant effects of early environmental conditions on mitochondrial density, metabolic capacity, or ROS production in either of the brain regions studied. These results were consistent when metabolic capacity and ROS production were standardised by mitochondrial density. Our findings suggest that metabolic function in the brain of _L. delicata_ is relatively resilient to early-life conditions. While we did not observe effects in the optic tecta and olfactory bulbs, evidence from other species indicates that the effects of CORT and temperature can be region-specific: incubation temperature alters metabolic capacity in some, but not all, brain nuclei of leopard geckos (_Eublepharis macularius_) [@coomber_independent_1997], and prenatal stress reduces mitochondrial efficiency in select nuclei of rats (_Rattus norvegicus_) [@gong_chronic_2011]. Thus, while our results point to robustness in the regions studied, other brain areas in _L. delicata_ may be more sensitive to prenatal CORT and temperature.

Alternatively, the effects of early environmental conditions on metabolic function and ROS production may be immediate rather than long-lasting. Mitochondria are highly responsive to environmental changes and energetic demands [@gyllenhammer2020developmental], and any alterations present during incubation may have dissipated once all animals developed under the same post-hatching environment. Nonetheless, even temporary changes in metabolic function during earlier stages would still be expected to affect oxidative damage, as these processes are cumulative and can lead to long-term consequences [@rice2002brain; @terman2006oxidative].

#### Oxidative stress brought about by experiencing early life CORT are region- and temperature- dependent
We found that DNA damage and lipid peroxidation were affected in region-specific ways. Early conditions did not significantly influence DNA damage in the olfactory bulbs, but they did alter lipid peroxidation in both brain nuclei studied. In the optic tecta, prenatal CORT reduced DNA damage in cold-incubated lizards but had no effect at higher incubation temperatures. In both regions, CORT also interacted with incubation temperature to influence lipid peroxidation, increasing it in CORT-treated cold-incubated lizards compared to the CORT-Hot group.

These contrasting patterns suggest that prenatal CORT and temperature may shape oxidative damage through molecule-specific protective pathways, potentially by upregulating different antioxidant systems. Glucocorticoids and temperature each appear to influence antioxidant systems in enzyme-specific ways rather than uniformly across pathways. For example, prenatal stress in rats increases Cu/Zn superoxide dismutase (SOD) and glutathione peroxidase (GPx) activity in the brain while catalase (CAT) was unaffected [@mcintosh1998glucocorticoids]. Similarly, in the soft-shelled turtle (_Pelodiscus sinensis_), CAT expression rises during arousal compared to hibernation, whereas GPx activity remains stable across these temperature-associated states [@tang2021antioxidant]. Such differential regulation could explain why CORT reduces DNA damage under cold incubation but reduces lipid peroxidation under warm incubation, if distinct antioxidant pathways are preferentially activated under different environmental conditions.

Alternative non-exclusive mechanisms could also contribute. CORT mobilises energy resources to facilitate physiological adjustments to stress [@sapolsky_how_2000]. Under cold incubation, when metabolic rates are expected to be lower, energy may be allocated towards antioxidant activity or DNA repair [@kawamura1991glucocorticoid; but see @mcintosh1998glucocorticoids]. In contrast, at warmer temperatures, increased metabolic demands may limit the resources available for such protective processes [see @kim2019carry]. These context-dependent changes may explain lower DNA damage in the optic tecta of CORT-Cold lizards, but they do not fully account for the reduced lipid peroxidation in CORT-Hot animals. Additional factors, such as temperature-dependent differences in membrane lipid composition [@zeis2019temperature], may further modulate molecule-specific sensitivity to oxidative stress. For example, lipid peroxidation levels have been shown to increase in cold-acclimated _Daphnia magna_ likely because of higher concentration of polyunsaturated fatty acids, which are more susceptible to oxidative damage [@zeis2019temperature]. Future studies should examine the effects of prenatal CORT and temperature on metabolic rate, antioxidant regulation, and molecule structure and composition in differen brain nuclei of reptiles to distinguish between these hypotheses.

#### Prenatal CORT exposure enhances chemical but not visual prey detection   
Detection latency decreased with CORT exposure when animals were presented with chemical, but not visual, stimuli. This effect was not related to alterations in mitochondrial physiology in the olfactory bulbs. Instead, prenatal CORT may influence other mechanisms involved in chemoreception. For instance, in rats (_R. norvegicus_), GCs like dexamethasone can enchance the ability to detect chemical stimuli by upregulating gene expression in the olfactory mucosa [@meunier2020olfactory]. Such molecular changes could alter chemosensory sensitivity in lizards, independent of mitochondrial function.

Alternatively, the observed differences in chemical perception may be driven by motivation. Prenatal CORT is expected to increase hunger through resource mobilization [@spencer2008post; @cossin-sevrin_effect_2022; but see @crino_corticosterone_2014-learn]. This could have led to increased motivation in CORT-treated lizards, stimulating a switch to a more active foraging mode, affecting the way they respond to chemical stimuli. For example, species of European lizards (family Lacertidae) that spend more time actively looking for food tend to rely more on chemoreception [@verwaijen2007relationships]. Although we fasted lizards and controlled for motivation in our analyses, it is possible that we did not fully capture factors related to the ecological relevance of the cues, the animals' internal state, or the interaction between the two. Further studies need to be conducted to understand how CORT can change foraging strategies in lizards and how this can affect their perception abilities.

@@

### Declaration of AI use
We declare Chat GPT was used for checking coding errors. All other parts of the manuscript were written by the authors and those parts where AI was used were checked and modified accordingly by the authors.

>>>>

Pablo Recio^1,2^,$\ddagger$ , Dalton C. Leibold^1^, Ondi L. Crino^1,2^, Christopher R. Friesen^3,4^, Daniel W.A. Noble^1^

@@

## Abstract
Cognitive processes such as the ability to perceive prey, are crucial for survival and reproduction. Early-life conditions like stress-related hormones or thermal environments, can shape cognitive abilities by influencing brain structure and function. Mitochondrial physiology is thought to be a central mechanism underlying these effects given that it is crucial for the energy production needed for brain function. Here, we investigated the combined influence of prenatal corticosterone (CORT)–a key stress hormone in reptiles–and incubation temperature on brain mitochondrial function and impacts on prey detection in the delicate skink (_Lampropholis delicata_). We manipulated egg CORT levels and incubation temperature, then assessed the ability of lizards to detect chemical and visual prey stimuli. Using flow cytometry, we measured metabolic function, reactive oxygen species (ROS) production, and oxidative stress in the olfactory bulbs and optic tecta-two brain regions involved in sensory processing. We predicted that CORT exposure and high temperatures would interact to shape perception of prey via changes in mitochondrial respiration and oxidative damage. While metabolic function remained robust to early life conditions, CORT and temperature interacted to influence oxidative stress in the optic tecta. Additionally, CORT-treated lizards responded faster when presented with chemical cues but not visual stimuli. Our findings suggest that prenatal conditions can have lasting effects on brain physiology and cognition, though the role of mitochondrial function in these processes remains unclear. Future research should explore alternative mechanisms that could link early-life stress to cell metabolism and cognitive performance.

@@

Cognition encompasses the ways in which animals acquire, process, and store information, enabling perception, learning, memory, and decision-making [@shettleworth]. It is essential for survival and reproduction, allowing individuals to adapt to changing environments [@dukas_evolutionary_2004]. However, cognitive abilities can vary considerably between individuals, with differences arising from genetic factors, environmental conditions, or a combination of both [@sakata_neural_2000; @dukas_evolutionary_2004]. The prenatal environment, in particular, plays a critical role in shaping brain development and cognition across a wide range of species [@sakata_neural_2000; @zhu_prenatal_2004; @crino_corticosterone_2014-learn; @amiel_effects_2017]. However, the impact of early-life environments has primarily been studied in the context of learning and memory [@zhu_prenatal_2004; @crino_corticosterone_2014-learn; @bebus_associative_2016; @abayarathna_effects_2020], neglecting how other important cognitive domains might be affected [but see @burger_antipredator_1998; @vila_pouca_quantity_2019]. For example, perception is an often overlooked, but fundamental cognitive process [@shettleworth]. Perceiving different stimuli is essential for behaviours critical to survival and reproductive succes such as locating food, avoiding predators, or interacting with conspecifics [@hairston1982fish; @pyke1984optimal; @burger_antipredator_1998; @desfilis2003stimulus; @recio2023conspecific]. Therefore, understanding how prenatal conditions affect the ability to detect prey can be important to understanding the broader consequences of early-life environments on fitness.

Prenatal environments influence cognitive abilities because the brain is particularly sensitive to environmental conditions during early stages of development [@zhu_prenatal_2004]. Early-life conditions can shape cognition by altering gene expression [@zhou2020effects], neurotransmitter production and neuroendocrine pathways [@amani2021perinatal], or brain structure and function [@amiel_effects_2017]. Additionally, developmental environments can have long-lasting effects on energy production and oxidative stress in the brain, which may be a key mechanism influencing cognitive function [@siegel1994basic; @du_dynamic_2009; @picard_mitochondria_2014; @picard_energetic_2018]. First, given the high energetic demands of cognitive processes [@mcnay_decreases_2000; @mann_coupling_2021; @alexandrov_neuronal_2022], cognitive performance is intrinsically linked to efficient mitochondrial respiration. Likewise, high cognitive abilities are typically associated with increased neuron density and functionality [@lefebvre_taxonomic_2011; @amiel_effects_2017], both of which can be impaired by excessive reactive oxygen species (ROS) production and oxidative stress [@finkel_oxidants_2000; @zhu_prenatal_2004; @du_dynamic_2009; @gong_chronic_2011; @hoffmann_mitochondrion_2018]. For instance, Hara et al. (2014) found that performance in visuospatial working-memory tasks is negatively correlated to the number of round-shaped mitochondria-associated with higher oxidative stress-per presynaptic terminal. Findings from different studies suggest that mitocondrial activity can be essential in shaping cognitive abilities. Thus, understanding how early-life conditions affect mitochondrial physiology may provide insights into the mechanisms underlying cognitive performance.

Mitochondria are highly responsive to environmental conditions during early development, with potential short-term and long-lasting consequences [reviewed in @gyllenhammer2020developmental]. For example, environmental conditions or hormone exposure can cause immediate disruptions to metabolic function, triggering cascading effects on oxidative stress that may persist later in life [reviewed in @gyllenhammer2020developmental]. Among the range of environmental factors that can shape mitochondrial physiology, some exert particularly strong and consistent effects. Stress-related hormones and temperature are particularly relevant for mitochondrial function due to their strong effects on metabolic processes [@sapolsky_how_2000; @stier2022experimental; @crino2024eggs]. Vertebrates respond to stressful situations through the activation of physiological processes including the production of  glucocorticoids (GCs), which are also important metabolic regulators [@sapolsky_how_2000]. In endothermic vertebrates, temperature plays a similar role in regulating metabolic activity [@stier2022experimental; @crino2024eggs]. In fact, heightened GC levels or changes in the thermal environment during early development have profound and sustained effects on mitochondrial respiration efficiency or oxidative stress [@zhu_prenatal_2004; @song2009prenatal; @costantini2011meta; @haussmann_embryonic_2012; @treidel2016temperature; @stier2022experimental; @crino2024eggs]. However, how the early environment influences mitochondrial physiology, and its associated cognitive consequences remains largely unknown outside of mammals [see @chaudhari2022early for a review]. In addition, whether and how these interactions affect perceptual abilities, such as prey detection, is still largely unexplored.

@@

#### Statistical analyses
We performed the analyses for each brain region/stimulus and each variable (mitochondrial density, metabolic capacity, metabolic capacity/mitochondrial density, ROS, ROS/mitochondrial density, DNA damage, lipid peroxidation, and detection latency) separately. We first fit a set of preliminary models for all the interest variables where we included the hormone (CORT versus Control), temperature (Cold versus Hot), and their interaction. These models also included the sex and age of the lizards at the time where the trials started (for detection latency) or when the lizards were euthanized (for all mitochondrial-related variables). For detection latency, we also included previous experience with the prey (familiar versus unknown), their performance on the motivation test [if they ate the cricket (1), or not (0)], and the interaction between CORT and motivation as fixed effects. We included the interaction between CORT and motivation because CORT can impact appetite [@conde2018stress]. All models included clutch identity as a random factor to account for clutch effects. For models of detection latency we also included a random effect of lizard identity because of our repeated measures design. The structure and results of these models are provided on Tables S1-S22 in _Supplementary Material_. After the preliminary models, we simplified models by excluding some of the factors that were not significant. However, we always included the main interest factors: CORT, temperature and their interaction. Random factors remain the same for the final models. All the response variables were mean centered and standardized by dividing by two times the standard deviation [@gelman2008scaling]. Before standardization, mitochondrial density, DNA damage, lipid peroxidation, and detection latency were log-transformed.

@@

_Optic tecta_: We did not find significant effects of CORT, temperature, or their interaction on mitochondrial density, metabolic capacity, or ROS production in the optic tecta, neither of CORT or incubation temperature on DNA damage or lipid peroxidation (see @fig-results_energy, @fig-results_oxidative, and Table S2). However, the interaction of CORT and temperature have significant effects on DNA damage ([(β~Control-Hot~ - β~CORT-Hot~) - (β~Control-Cold~ - β~CORT-Cold~)]: mean = `r format_dec((mean(DNA_OT$Control_Hot) - mean(DNA_OT$CORT_Hot)) - (mean(DNA_OT$Control_Cold) - mean(DNA_OT$CORT_Cold)), 3)`, pMCMC `r format_p(pmcmc((DNA_OT$Control_Hot - DNA_OT$CORT_Hot) - (DNA_OT$Control_Cold - DNA_OT$CORT_Cold)), 3, equal = TRUE)`), and lipid peroxidation ([(β~Control-Hot~ - β~CORT-Hot~) - (β~Control-Cold~ - β~CORT-Cold~)]: mean = `r format_dec((mean(LP_OT$Control_Hot) - mean(LP_OT$CORT_Hot)) - (mean(LP_OT$Control_Cold) - mean(LP_OT$CORT_Cold)), 3)`, pMCMC `r format_p(pmcmc((LP_OT$Control_Hot - LP_OT$CORT_Hot) - (LP_OT$Control_Cold - LP_OT$CORT_Cold)), 3, equal = TRUE)`) in the optic tecta. CORT elevations had no effect on hot-incubated lizards (β~Control-Hot~ - β~CORT-Hot~: mean = `r format_dec(mean(DNA_OT$Control_Hot) - mean(DNA_OT$CORT_Hot),3)`, pMCMC `r format_p(pmcmc(DNA_OT$Control_Hot - DNA_OT$CORT_Hot), 3, equal = TRUE)`), but it decreased total DNA damage in cold-incubated animals (β~Control-Cold~ - β~CORT-Cold~: mean = `r format_dec(mean(DNA_OT$Control_Cold) - mean(DNA_OT$CORT_Cold),3)`, pMCMC `r format_p(pmcmc(DNA_OT$Control_Cold - DNA_OT$CORT_Cold), 3, equal = TRUE)`) (see @fig-results_oxidative C, E). The direction of the effects of CORT on lipid peroxidation were the opposite on eggs incubated at high temperatures (β~Control-Hot~ - β~CORT-Hot~: mean = `r format_dec(mean(LP_OT$Control_Hot) - mean(LP_OT$CORT_Hot),3)`, pMCMC `r format_p(pmcmc(LP_OT$Control_Hot - LP_OT$CORT_Hot), 3, equal = TRUE)`) than at low temperatures (β~Control-Cold~ - β~CORT-Cold~: mean = `r format_dec(mean(LP_OT$Control_Cold) - mean(LP_OT$CORT_Cold),3)`, pMCMC `r format_p(pmcmc(LP_OT$Control_Cold - LP_OT$CORT_Cold), 3, equal = TRUE)`), with significant differences between CORT-Hot and CORT-Cold lizards (β~CORT-Hot~ - β~CORT-Cold~: mean = `r format_dec(mean(LP_OT$CORT_Hot) - mean(LP_OT$CORT_Cold),3)`, pMCMC `r format_p(pmcmc(LP_OT$CORT_Hot - LP_OT$CORT_Cold), 3, equal = TRUE)`; see @fig-results_oxidative C). In addition, we found that females had higher DNA damage levels than males, and lipid peroxidation increased with age (see Table S4).

@@

_Chemical perception (Olfactory bulbs)_: We found a positive effect of mitochondrial density on ROS production (mean β~Mit density~ ~-~ ~ROS~ = `r format_dec(mean(OB_coeff_mitodensity_ros), 3)`, pMCMC `r format_p(pmcmc(OB_coeff_mitodensity_ros), 3, equal = TRUE)`). However, there were no significant relationships between mitochondrial function or oxidative stress on prey detection, nor between other any other variables (see @fig-sem_results_OB and Table S5).

@@

## Discussion
We examined the impact of prenatal temperature and CORT on visual and chemical perception to test whether metabolic function underpinned perception in lizards, as previously shown in mammals [@hara_presynaptic_2014; @chaudhari2022early]. Overall, metabolic processes appeared largely resilient to prenatal conditions. However, early-life conditions produced region- and context-dependent effects on oxidative damage and prey detection. We observed interactive effects of prenatal CORT and incubation temperature on lipid peroxidation in the two brain regions examined, while the interaction affected DNA damage only in the optic tecta. Behaviourally, prenatal CORT exposure enhanced detection of chemical but not visual prey stimuli. Yet, we found no evidence that metabolism explained behavioural variation in prey detection behaviours.

#### Metabolic processes in multiple brain regions are robust to early life conditions
Contrary to our predictions, we found no significant effects of early environmental conditions on mitochondrial density, metabolic capacity, or ROS production in the optic tecta or the olfactory bulbs. These results were consistent when metabolic capacity and ROS production were standardised by mitochondrial density. Our findings suggest that metabolic function in the brain of _L. delicata_ is relatively resilient to early-life conditions. While we did not observe effects in the optic tecta and olfactory bulbs, evidence from other species indicates that the effects of prenatal CORT and temperature can be region-specific. For example, incubation temperature alters metabolic capacity in some, but not all, brain nuclei of leopard geckos (_Eublepharis macularius_) [@coomber_independent_1997], and prenatal stress reduces mitochondrial efficiency in select nuclei of rats (_Rattus norvegicus_) [@gong_chronic_2011]. Thus, while our results point to robustness in the regions studied, other brain areas in _L. delicata_ may be more sensitive to prenatal CORT and temperature.

Alternatively, the effects of early environmental conditions on metabolic function and ROS production may be present during incubation, but diminish or dissapear over time. Mitochondria are highly responsive to environmental changes and energetic demands [@gyllenhammer2020developmental], and any alterations present during incubation may have dissipated once all animals developed under the same post-hatching environment. In our experiment, lizards were housed in standardized conditions prior to euthanasia at 238-356 days post-hatching (mean = 273.25). It is possible that the effects of prenatal CORT and temperature exposure had lessened by the time metabolic function and ROS production were measured. Nonetheless, even temporary changes in metabolic function during earlier stages would still be expected to affect oxidative damage, as these processes are cumulative and can lead to long-term consequences [@rice2002brain; @terman2006oxidative].

#### Oxidative stress brought about by experiencing early life CORT are region- and temperature- dependent
We found that DNA damage and lipid peroxidation were affected in region-specific ways. Early conditions did not significantly influence DNA damage in the olfactory bulbs, but they did alter lipid peroxidation in the olfactory bulbs and the optic tecta. In the optic tecta, prenatal CORT reduced DNA damage in cold-incubated lizards but had no effect in hot-incubated animals. In both regions, CORT also interacted with incubation temperature to influence lipid peroxidation. Lizards treated with prenatal CORT and incubated at cold temperatures had higher levels of lipid peroxidation compared to the lizards from the CORT-Hot group. These contrasting patterns suggest that prenatal CORT and temperature may shape oxidative damage through molecule-specific protective pathways, potentially by upregulating different antioxidant systems. For example, prenatal stress in rats increases Cu/Zn superoxide dismutase (SOD) and glutathione peroxidase (GPx) activity in the brain while catalase (CAT) was unaffected [@mcintosh1998glucocorticoids]. Similarly, in the soft-shelled turtle (_Pelodiscus sinensis_), CAT expression elevates with increasing temperatures as turtles emerge from hibernation, whereas GPx activity remains stable [@tang2021antioxidant]. Such differential regulation of antioxidant systems could explain why CORT reduces DNA damage under cold incubation but reduces lipid peroxidation under hot incubation.

However, other mechanisms could be contributing to our results. For example, CORT mobilises energy resources to facilitate physiological adjustments to stress and environmental challenges [@sapolsky_how_2000]. Under cold incubation, when metabolic rates are expected to be lower, energy may be allocated towards antioxidant activity or DNA repair [@kawamura1991glucocorticoid; but see @mcintosh1998glucocorticoids]. In contrast, at warmer temperatures, increased metabolic demands may limit the resources available for such protective processes [see @kim2019carry]. These context-dependent changes may explain lower DNA damage in the optic tecta of CORT-Cold lizards, but they do not fully account for the reduced lipid peroxidation in CORT-Hot animals. Additional factors, such as temperature-dependent differences in membrane lipid composition [@zeis2019temperature], may further modulate molecule-specific sensitivity to oxidative stress. For example, lipid peroxidation levels have been shown to increase in cold-acclimated _Daphnia magna_ likely because of higher concentration of polyunsaturated fatty acids, which are more susceptible to oxidative damage [@zeis2019temperature]. Future studies should examine the effects of prenatal CORT and temperature on metabolic rate, antioxidant regulation, and molecule structure and composition in differen brain nuclei of reptiles to parse between these hypotheses.

#### Prenatal CORT exposure enhances chemical but not visual prey detection   
Detection latency decreased with CORT exposure when animals were presented with chemical, but not visual, stimuli. This effect was not related to alterations in mitochondrial physiology in the olfactory bulbs. Instead, prenatal CORT could influence other mechanisms involved in chemoreception. For instance, in rats (_R. norvegicus_), dexamethsone (a synthetic GC) enhances the ability to detect chemical stimuli by upregulating gene expression in the olfactory mucosa [@meunier2020olfactory]. If similar GC-dependent transcriptional changes occur in lizards,they could alter chemosensory sensitivity independent of mitochondrial function.

Alternatively, the observed differences in chemical perception may be driven by treatment effects on motivation. We fasted lizards in our experiment for two days prior to behavioural assays to standardize hunger levels. However, the experimental treatments could have affected factors related to the animals’ internal state that we did not measure that influenced motivation. For example, prenatal CORT can increase hunger through resource mobilisation [@spencer2008post; @cossin-sevrin_effect_2022; but see @crino_corticosterone_2014-learn]. This could have led to increased motivation in CORT-treated lizards decreasing their time to respond to prey chemical stimuli. Furthermore, the absence of CORT-related effects on visual perception could occur because actively foraging lizards often prioritize chemosensory cues [@verwaijen2007relationships], so changes in motivation may selectively affect chemical perception. Further studies need to be conducted to understand how CORT can change foraging strategies in lizards and how this can affect their perception abilities.

@@

### Declaration of AI use
We declare Chat GPT was used for inspecting the code for errors when necessary. All other parts of the manuscript were written by the authors.
